Space for WHAM: a multi-region, multi-stock generalization of the Woods Hole Assessment Model with an application to black sea bass

### 2025-03-18 paper/paper.tex
add edits from Chris and a little tidying of the repo

  pdfauthor={Timothy J. Miller1; Kierstin Curti; Alexander C. Hansell},

@@

\title{Space for WHAM: a multi-region, multi-stock generalization of the Woods Hole Assessment Model with an application to black sea bass}
\author{Timothy J. Miller\textsuperscript{1} \and Kierstin Curti \and Alexander C. Hansell}
\date{04 March, 2025}

@@

\hypertarget{main-message}{%
\subsection*{Main Message}\label{main-message}}
\addcontentsline{toc}{subsection}{Main Message}

Describes the multi-region, multi-stock generalization of WHAM and its usage to evaluate evidence of alternative hypotheses about temperature effects on recruitment and age 1 natural mortality of black sea bass stock components.

@@

A state-space age-structured stock assessment models can be used to estiamte time and age-varying population attributes as random effects using maximum marginal likelihood or Bayesian fitting procedures (Nielsen and Berg 2014; Cadigan 2016; Miller et al. 2016b). This estimation approach is considered an essential feature of gold-standard assessment models that we use in tactical management of commercially important fish stocks (Punt et al. 2020). The State-space Assessment Model (SAM, Nielsen and Berg 2014) continues to be developed and remains widely used within the International Council for the Exploration of the Sea (ICES) to assess European fish stocks. Various state-space models are being used to manage cod and plaice stocks in the waters of Eastern Canada (Perreault et al. 2020; Varkey et al. 2022) and to the south, the Woods Hole Assessment Model (WHAM, Miller and Stock 2020; Stock and Miller 2021) is now used to assess many fish stocks in the Northwest Atlantic Ocean (NEFSC 2022a, 2022b; NEFSC 2024).

@@

We developed this extension of WHAM in concert with a research track assessment process for black sea bass (\textit{Centropristis striata}) (NEFSC 2023). Previous black sea bass assessments assumed two stock components (North and South), each assessed with a separate statistical catch at age assessment. The stock was split into two regions to account for spatial dynamics and improve model diagnostics (ASMFC 2016). Additionally, black sea bass have complex spatial dynamics and make seasonal migrations, moving inshore in the spring and offshore in the fall (Moser and Shepherd 2009). Black sea bass distribution has been linked to warming waters on the northwest shelf (Bell et al. 2015) and it is hypothesized that the distribution and productivity of the black sea bass stock is especially susceptible to long-term changes in temperature causing increases in abundance further north (Hare 2016). Thus, a focus of the research track assessment was to develop an assessment platform that could capture multiple stock components, complex seasonal movement and explore environmental drivers of population dynamics (NEFSC 2023).To illustrate the usage of WHAM 2.0 we apply it to two stock components of black sea bass off the coast of the NEUS using most of the model assumptions that were accepted during the peer-review process. However, here we evaluate evidence for alternative hypotheses of temporal variation and effects specifically of bottom temperature on recruitment and natural mortality of age 1 individuals.

When modeling multiple stock components occurring in different regions further dimensionality is added to the class of possible equilbrium reference points (e.g., Kapur et al. 2021). Short term projections are also an important part of the management process and the separation of observation and process errors in state-space models allows uncertainty in projected attributes such as recruitment and spawning stock biomass (SSB) without external simulation exercises. Finally, the inclusion of explicit effects of environmental covariates in the model also requires some assumptions on how they are treated in projections and reference points. We demonstrate some of options of WHAM version 2.0 to provide biological reference points and short-term projections.

@@

Because individuals may be alive in one of several regions or harvested in one of several fleets, it is helpful to consider these as distinct categories or states and treat the number of individuals occurring in each category over time as a multi-state model. Approaches to modeling transitions among these states may treat time discretely (e.g., Arnason 1972; Schwarz et al. 1993) or continuously (e.g., Hearn et al. 1987; Commenges 1999; Andersen and Keiding 2002). We can define a probability transition matrix (PTM) that describes the probability of individuals occurring in different states at the end of a time interval \(\delta\) (columns), conditional on being in each of those state at beginning of the interval (rows). For fish populations, these states would be defined as being alive in a particular region or deceased due to fishing from a particular fleet or natural mortality. The time interval \(i\) with duration \(\delta_i\) would be a season and the PTMs would be uniquely defined for each stock by season \(i\), year \(y\), and age \(a\) on January 1. Each row and column of the PTM correspond to one the states: alive in region \(r\), dead in fleet \(f\), or dead from natural causes. The probabilities in each row sum to unity (must be in one of the states) and assume an individual is in the state that corresponds to that row at the beginning of the interval. Given \(n_R\) regions and \(n_F\) fleets, the square PTM (\(n_R + n_F + 1\) rows and columns) as a function of sub-matrices is
\begin{equation}\label{eq:ptm}
  \mathbf{P}_{y,a,i} = \begin{bmatrix}
    \mathbf{O}_{y,a,i} & \mathbf{H}_{y,a,i} & \mathbf{D}_{y,a,i} \\
    0 & \mathbf{I}_{H} & 0\\
    0 & 0 & 1
  \end{bmatrix}
\end{equation}
where
\begin{equation*}
  \mathbf{O}_{y,a,i} = 
  \begin{bmatrix}
    O_{y,a,i}(1,1) & \cdots & O_{y,a,i}(1,n_R) \\
    \vdots & \ddots & \vdots \\
    O_{y,a,i}(n_R,1) & \cdots & O_{y,a,i}(n_R,n_R)
  \end{bmatrix}
\end{equation*}
is the \(n_R \times n_R\) matrix defining survival and occurring in each region at the end of the interval,
\begin{equation*} 
  \mathbf{H}_{y,a,i} = 
  \begin{bmatrix}
    H_{y,a,i}(1,1) & \cdots & H_{y,a,i}(1,n_F) \\
    \vdots & \ddots & \vdots \\
    H_{y,a,i}(n_R,1) & \cdots & H_{y,a,i}(n_R,n_F)
  \end{bmatrix}
\end{equation*}
is the \(n_R \times n_F\) matrix defining probabilities of being captured in each fleet during the interval, and \(\mathbf{D}_{y,a,i}\) is the \(n_R x 1\) matrix of probabilities of dying due to natural mortality during the interval. We have the identity matrix \(\mathbf{I}_{H}\) for the states of capture by each fleet and a 1 for the natural mortality state because the probabilities of being in one of the mortality states given starting the interval in that state is unity (no zombies allowed).

@@

When there is more than one region, WHAM can model survival and movement as processes occurring sequentially or simultaneously. The sequential assumption is used widely in spatially explicit model (e.g., Stock Synthesis, Methot and Wetzel 2013). Under the sequential assumption, survival and death occur over the interval and movement among regions occurs instantly at either the beginning or the end of the interval. WHAM is configured to have movement occur after survival:
\begin{equation*}
  \mathbf{O}_{y,a,i} = \mathbf{S}_{y,a,i}\boldsymbol{\mu}_{y,a,i}
\end{equation*}
where \(\mathbf{S}_{y,a,i}\) is a \(n_R \times n_R\) diagonal matrix of proportions surviving in each region (given they start in that region)
\begin{equation*}
\mathbf{S}_{y,a,i} = 
  \begin{bmatrix}
    e^{-Z_{y,a,i,1}\delta_i} & 0 & \cdots & 0 \\
    0 & e^{-Z_{y,a,i,2}\delta_i} & \cdots & 0 \\
    \vdots & \vdots & \ddots & \vdots \\
    0 & \cdots & 0 & e^{-Z_{y,a,i,n_R}\delta_i}
  \end{bmatrix}
\end{equation*}
and \(\boldsymbol{\mu}_{y,a,i}\) is a \(n_R \times n_R\) matrix of probabilities of moving from one region to another or staying in the region they occurred at the beginning of the interval:
\begin{equation*}
\boldsymbol{\mu}_{y,a,i} = 
  \begin{bmatrix}
    1-\sum_{r' \neq 1} \mu_{1\rightarrow r',y,a,i} & \mu_{1\rightarrow 2,y,a,i} & \cdots & \mu_{1\rightarrow R,y,a,i} \\
    \mu_{2\rightarrow 1,y,a,i} & 1-\sum_{r' \neq 2} \mu_{2\rightarrow r',y,a,i} & \cdots & \mu_{2\rightarrow R,y,a,i} \\
    \vdots & \vdots & \ddots & \vdots \\
    \mu_{R\rightarrow 1,y,a,i} & \cdots & \mu_{R\rightarrow R-1,y,a,i} & 1-\sum_{r' \neq R} \mu_{R\rightarrow r',y,a,i}
  \end{bmatrix}
\end{equation*}

@@

When survival and movement are assumed to occur simultaneously, all movement and mortality parameters are instantaneous rates. We obtain the probability transition matrix over an interval \(\delta_i\) by exponentiating the instantaneous rate matrix (Miller and Andersen 2008)
\begin{equation*}
\mathbf{P}_{y,a,i} = e^{\mathbf{A}_{y,a,i}\delta_i}
\end{equation*}
The instantaneous rate matrix takes rates of movement between regions and the mortality rates for each fleet and region. Along the diagonal is the negative of the sum of the other rates (the hazard) so each row sums to zero. For two regions and one fleet operating in each region:
\begin{equation*}
 \mathbf{A}_{y,a,i} = \begin{bmatrix}
 a_{y,a,i,1} & \mu_{1\rightarrow 2,y,a,i} & F_{y,a,i,1} & 0 & M_{y,a,1} \\
 \mu_{2\rightarrow 1,y,a,i} &  a_{y,a,i,2} & 0 & F_{y,a,i,2} & M_{y,a,2} \\
 0 & 0 & 0 & 0 & 0 \\
 0 & 0 & 0 & 0 & 0 \\
 0 & 0 & 0 & 0 & 0
 \end{bmatrix}.
\end{equation*}
where \(a_{y,a,i,r} = -(\mu_{r\rightarrow r',y,a,i} + F_{y,a,i,r_f} + M_{y,a,r})\). When there is one region, \(n_f\) fleets, and \(\delta_i = 1\), exponentiating the instantaneous rate matrix results in the PTM defined in Eq. \ref{eq:ptm_1_region}.

@@

When there are \(n_R\) regions and \(n_F\) fleets, the vector of abundance in each state at age \(a>1\) on January 1 is \(\mathbf{N}_{y,a} = (\mathbf{N}_{O,y,a}', \mathbf{0}')'\) where \(\mathbf{N}_{O,y,a} = (N_{y,a,1}, \ldots, N_{y,a,n_R})'\) is the number in the states corresponding to being alive in each region and \(\mathbf{0}\) is a vector (\(n_F+1\)) for the numbers captured in each fleet and dead from natural mortality because no age \(a\) fish have died yet on January 1.

Each stock \(s\) is assumed to spawn and recruit in one region \(r_s\). So for age \(a=1\), \(\mathbf{N}_{O,y,1}\) is 0 except for row \(r = r_s\). Options for configuring recruitment (\(N_{y,1,r_s}\)) for each stock are the same as previous versions of WHAM (Miller and Stock 2020). If recruitment is assumed to be a function of SSB, it is only the spawning population in region \(r_s\) at the time of spawning that constitutes the SSB in the stock-recruit function. However, models can configure spawning individuals to occur in other regions at the time of spawning. Aside from treating recruitment as a random walk, the general model for annual recruitment as random effects is
\begin{equation*}
\log\left(N_{y,1,r_s}\right)|\text{SSB}_{y-1,r_s} =  f\left(\text{SSB}_{y-1,r_s}\right) + \varepsilon_{y,1,r_s}
\end{equation*}
where\\
\[\text{SSB}_{y,r_s} =  \sum^A_{a=1}  w_{y,a} \text{mat}_{y,a} \mathbf{O}_{y,a,r_s}(t_s)' \mathbf{N}_{O,y,a}\]
and \(w_{y,a}\) is the mean weight at age of spawning individuals, \(\text{mat}_{y,a}\) is the maturity at age, and \(\mathbf{O}_{y,a,r_s}(t_s)\) are the probabilities of surviving and occurring in region \(r_s\) at time \(t_s\) given being alive in each region at the start of the year (the \(r_s\) column of the upper-left submatrix of Eq. \ref{eq:ptm_spawn}).

As in previous versions of WHAM, the transitions in numbers at age from one year to another after recruitment can be treated deterministically or as functions of random effects. The predicted numbers at age in year \(y\) at age \(a\) for a given stock are vector analogs (\(\mathbf{N}_{O,y,a}\)) of the equations for numbers at age in the standard WHAM model (Stock and Miller 2021). For ages \(a = 2,\ldots, A-1\), where \(A\) is the plus group, the expected number alive in each region at the beginning of the following year and next age class age can be obtained from the first \(n_R\) elements of the vector
\[\mathbf{P}_{y-1,a-1}' \mathbf{N}_{y-1,a-1}.\]
The numbers alive in each region can also be modeled more simply using the sub-matrix \(\mathbf{O}_{y,a}\). The general model for the transitions in abundance at age is
\begin{equation*}
\log\left(\mathbf{N}_{O,y,a}\right)|\mathbf{N}_{O,y-1,a-1} =  \log\left(\mathbf{O}_{y-1,a-1}' \mathbf{N}_{O,y-1,a-1}\right) + \boldsymbol{\varepsilon}_{y,a}
\end{equation*}
for ages \(a = 2,\ldots, A-1\), and for the plus group
\begin{equation*}
\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,a-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}' \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}' \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A}
\end{equation*}
where \(\boldsymbol{\varepsilon}_{y,a}\) is the vector of region-specific errors for a given stock, which are independent across stocks and regions, but the same correlation structures as previous versions are possible across ages and years for a given stock and region.
When there is autocorrelation with age, WHAM now assumes this applies only to ages \(a>1\) by default so that recruitment random effects are independent of those for for the annual transitions of older age classes. So the general covariance structure for a given stock at ages \(a>1\) in region \(r\) is that of a two-dimensional first-order autoregressive (2DAR1) process
\begin{equation*}
  Cov\left(\varepsilon_{y,a,r},\varepsilon_{y',a',r}\right) =   \frac{\rho_{N,\text{age},r}^{|a-a'|}\rho_{N,\text{year},r}^{|y-y'|}\sigma_{N,a,r}\sigma_{N,a',r}}{\left(1 -  \rho_{N,\text{age},r}^2\right)\left(1 - \rho_{N,\text{year},r}^2\right)} 
\end{equation*}
and that for age 1 is just AR1 across years
\begin{equation*}
  Cov\left(\varepsilon_{y,1},\varepsilon_{y',1}\right) =   \frac{\rho_{N,1,\text{year}}^{|y-y'|}\sigma^2_{N,1}}{1 - \rho_{N,1,\text{year}}^2}.
\end{equation*}
When the transitions in abundance at age are treated deterministically, \(\boldsymbol{\varepsilon}_{y,a} = 0\). Since recruitment for a given stock currently only occurs in one region \(r_s\) there is only a single vector of time-varying recruitment random effects for each stock.

@@

Initial numbers at age for each stock and region can be treated as age-specific fixed effects or with an equilibrium assumption as in previous versions of WHAM. For the equilibrium option there are two parameters for each stock: the stock-specific fully-selected fishing mortality rate \(\log \widetilde{F}\) and the recruitment in year 1 \(\log N_{1,1,r_s}\). A stock-specific equilibrium fishing mortality at age by fleet \({\tilde F}_{a,f}\) is the product of \(\widetilde{F}\) and the selectivity across fleets in the first year
\begin{equation*}
  \text{sel}_{1,a,f} = \frac{F_{1,a,f}}{\max_a \sum_{f=1}^{n_F} F_{1,a,f}}.
\end{equation*}
We use \(\widetilde{F}_{a,f}\) to define an equilibrium abundance per recruit by region at age \(a\) conditional on recruiting to each region
\begin{equation}\label{eq:npraa}
 \widetilde{\mathbf{O}}_{a} = \left\{
 \begin{array}{ll}
\prod^{a-1}_{j=0}\mathbf{O}_{j}  & 1\leq a<A\\
\left[\prod^{a-1}_{j=0}\mathbf{O}_{j}\right] \left(\mathbf{I} - \mathbf{O}_{A}\right)^{-1} & a = A
 \end{array}
\right.
\end{equation}
where \(\mathbf{O}_{j}\) is the equilibrium probability of surviving age \(a\) and occurring in each region and \(\mathbf{O}_{0} = \mathbf{I}\). Natural mortality and movement rates in the first year of the model are also used in Eq. \ref{eq:npraa}. For the plus group \(a=A\), \(\left(\mathbf{I} - \mathbf{O}_{A}\right)^{-1}\) is a ``fundamental matrix'\,' derived using the matrix version of the geometric series (Kemeny and Snell 1960). Recall that recruitment for stock \(s\) only occur in region \(r_s\) so, the equilibrium initial numbers at age \(a\) by region are
\[\mathbf{N}_{O,1,a} = \widetilde{\mathbf{O}}_{a}' \mathbf{N}_{O,1,1}.\]

@@

When treated as constant parameters, natural mortality rates may be stock-, region-, and age-specific. When random effects are used, the same 2DAR1 structure with age and year as described by Stock and Miller (2021) can be configured for a given stock and region and any environmental covariate effects can be stock-, region-, and age-specific. So the general model for natural mortality is
\begin{equation*}
  \log M_{y,a,r} = \theta_{M,r} + \varepsilon_{M,r,y,a} + \sum^{n_E}_{k=1} \beta_{M,r,a,k} E_{k,y}.
\end{equation*}
The general covariance structure for random effects are modeled most generally as 2DAR1 random effect with age and(or) year where the covariance is
\begin{equation*}
  Cov\left(\varepsilon_{M,y,a,r},\varepsilon_{M,y',a',r}\right) =   \frac{\rho_{M,\text{age},r}^{|a-a'|}\rho_{M,\text{year},r}^{|y-y'|}\sigma^2_{M,r}}{\left(1 -  \rho_{M,\text{age},r}^2\right)\left(1 - \rho_{M,\text{year},r}^2\right)}.
\end{equation*}

@@

>>>>

  pdfauthor={Timothy J. Miller1; Kiersten Curti; Alexander C. Hansell},

@@

\title{Space for WHAM: a multi-region, multi-stock generalization of the Woods Hole Assessment Model with an application to black sea bass}
\author{Timothy J. Miller\textsuperscript{1} \and Kiersten Curti \and Alexander C. Hansell}
\date{18 March, 2025}

@@

\textbf{keywords:} stock assessment, state-space, multi-stock, multi-region, age-structured, black sea bass

@@

A state-space age-structured stock assessment model can be used to estiamte time and age-varying population attributes as random effects using maximum marginal likelihood or Bayesian fitting procedures (Nielsen and Berg 2014; Cadigan 2016; Miller et al. 2016b). This estimation approach is considered an essential feature of gold-standard assessment models that we use in tactical management of commercially important fish stocks (Punt et al. 2020). The State-space Assessment Model (SAM, Nielsen and Berg 2014) continues to be developed and remains widely used within the International Council for the Exploration of the Sea (ICES) to assess European fish stocks. Various state-space models are being used to manage cod and plaice stocks in the waters of Eastern Canada (Perreault et al. 2020; Varkey et al. 2022) and to the south, the Woods Hole Assessment Model (WHAM, Miller and Stock 2020; Stock and Miller 2021) is now used to assess many fish stocks in the Northwest Atlantic Ocean (NEFSC 2022a, 2022b; NEFSC 2024).

@@

We developed this extension of WHAM in concert with a research track assessment process for black sea bass (\textit{Centropristis striata}) (NEFSC 2023). Previous black sea bass assessments assumed two stock components (North and South), each assessed with a separate statistical catch at age assessment. The stock was split into two regions to account for spatial dynamics and improve model diagnostics (ASMFC 2016). Additionally, black sea bass have complex spatial dynamics and make seasonal migrations, moving inshore in the spring and offshore in the fall (Moser and Shepherd 2009). Black sea bass distribution has been linked to warming waters on the northwest shelf (Bell et al. 2015) and it is hypothesized that the distribution and productivity of the black sea bass stock is especially susceptible to long-term changes in temperature causing increases in abundance further north (Hare et al. 2016). Thus, a focus of the black sea bass research track assessment was to develop an assessment platform that could capture multiple stock components, complex seasonal movement and explore environmental drivers of population dynamics (NEFSC 2023).To illustrate the usage of WHAM 2.0 we apply it to two stock components of black sea bass off the coast of the Northeast United States (NEUS) using most of the model assumptions that were accepted during the peer-review process. However, here we evaluate evidence for alternative hypotheses of temporal variation and effects specifically of bottom temperature on recruitment and natural mortality of age 1 individuals.

When modeling multiple stock components occurring in different regions further dimensionality is added to the class of possible equilbrium reference points (e.g., Kapur et al. 2021). Short term projections are also an important part of the management process and the separation of observation and process errors in state-space models allows uncertainty in projected attributes such as recruitment and spawning stock biomass (SSB) without external simulation exercises. Finally, the inclusion of explicit effects of environmental covariates in the model also requires some assumptions on how they are treated in projections and reference points. We demonstrate some of the options of WHAM version 2.0 to provide biological reference points and short-term projections.

@@

Because individuals may be alive in one of several regions or harvested in one of several fleets, it is helpful to consider these as distinct categories or states and treat the number of individuals occurring in each category over time as a multi-state model. Approaches to modeling transitions among these states may treat time discretely (e.g., Arnason 1972; Schwarz et al. 1993) or continuously (e.g., Hearn et al. 1987; Commenges 1999; Andersen and Keiding 2002). We can define a probability transition matrix (PTM) that describes the probability of individuals occurring in different states at the end of a time interval \(\delta\) (columns), conditional on being in each of those state at beginning of the interval (rows). For fish populations, these states would be defined as being alive in a particular region or deceased due to fishing from a particular fleet or natural mortality. The time interval \(i\) with duration \(\delta_i\) would be a season and the PTMs would be uniquely defined for each stock by year \(y\), age \(a\), and season \(i\) on January 1. Each row and column of the PTM correspond to one of the states: alive in region \(r\), dead in fleet \(f\), or dead from natural causes. The probabilities in each row sum to unity (must be in one of the states) and assume an individual is in the state that corresponds to that row at the beginning of the interval. Given \(n_R\) regions and \(n_F\) fleets, the square PTM (\(n_R + n_F + 1\) rows and columns) as a function of sub-matrices is
\begin{equation}\label{eq:ptm}
  \mathbf{P}_{y,a,i} = \begin{bmatrix}
    \mathbf{O}_{y,a,i} & \mathbf{H}_{y,a,i} & \mathbf{D}_{y,a,i} \\
    0 & \mathbf{I}_{H} & 0\\
    0 & 0 & 1
  \end{bmatrix}
\end{equation}
where
\begin{equation*}
  \mathbf{O}_{y,a,i} = 
  \begin{bmatrix}
    O_{y,a,i}(1,1) & \cdots & O_{y,a,i}(1,n_R) \\
    \vdots & \ddots & \vdots \\
    O_{y,a,i}(n_R,1) & \cdots & O_{y,a,i}(n_R,n_R)
  \end{bmatrix}
\end{equation*}
is the \(n_R \times n_R\) matrix defining survival and occurring in each region at the end of the interval,
\begin{equation*} 
  \mathbf{H}_{y,a,i} = 
  \begin{bmatrix}
    H_{y,a,i}(1,1) & \cdots & H_{y,a,i}(1,n_F) \\
    \vdots & \ddots & \vdots \\
    H_{y,a,i}(n_R,1) & \cdots & H_{y,a,i}(n_R,n_F)
  \end{bmatrix}
\end{equation*}
is the \(n_R \times n_F\) matrix defining probabilities of being captured in each fleet during the interval, and \(\mathbf{D}_{y,a,i}\) is the \(n_R x 1\) matrix of probabilities of dying due to natural mortality during the interval. We have the identity matrix \(\mathbf{I}_{H}\) for the states of capture by each fleet and a 1 for the natural mortality state because the probabilities of being in one of the mortality states given starting the interval in that state is unity (no zombies allowed).

@@

When there is more than one region, WHAM can model survival and movement as processes occurring sequentially or simultaneously. The sequential assumption is used widely in spatially explicit models (e.g., Stock Synthesis, Methot and Wetzel 2013). Under the sequential assumption, survival and death occur over the interval and movement among regions occurs instantly at either the beginning or the end of the interval. WHAM is configured to have movement occur after survival:
\begin{equation*}
  \mathbf{O}_{y,a,i} = \mathbf{S}_{y,a,i}\boldsymbol{\mu}_{y,a,i}
\end{equation*}
where \(\mathbf{S}_{y,a,i}\) is a \(n_R \times n_R\) diagonal matrix of proportions surviving in each region (given they start in that region)
\begin{equation*}
\mathbf{S}_{y,a,i} = 
  \begin{bmatrix}
    e^{-Z_{y,a,i,1}\delta_i} & 0 & \cdots & 0 \\
    0 & e^{-Z_{y,a,i,2}\delta_i} & \cdots & 0 \\
    \vdots & \vdots & \ddots & \vdots \\
    0 & \cdots & 0 & e^{-Z_{y,a,i,n_R}\delta_i}
  \end{bmatrix}
\end{equation*}
and \(\boldsymbol{\mu}_{y,a,i}\) is a \(n_R \times n_R\) matrix of probabilities of moving from one region to another or staying in the region they occurred at the beginning of the interval:
\begin{equation*}
\boldsymbol{\mu}_{y,a,i} = 
  \begin{bmatrix}
    1-\sum_{r' \neq 1} \mu_{1\rightarrow r',y,a,i} & \mu_{1\rightarrow 2,y,a,i} & \cdots & \mu_{1\rightarrow R,y,a,i} \\
    \mu_{2\rightarrow 1,y,a,i} & 1-\sum_{r' \neq 2} \mu_{2\rightarrow r',y,a,i} & \cdots & \mu_{2\rightarrow R,y,a,i} \\
    \vdots & \vdots & \ddots & \vdots \\
    \mu_{R\rightarrow 1,y,a,i} & \cdots & \mu_{R\rightarrow R-1,y,a,i} & 1-\sum_{r' \neq R} \mu_{R\rightarrow r',y,a,i}
  \end{bmatrix}
\end{equation*}

@@

When survival and movement are assumed to occur simultaneously, all movement and mortality parameters are instantaneous rates. We obtain the PTM over an interval \(\delta_i\) by exponentiating the instantaneous rate matrix (Miller and Andersen 2008)
\begin{equation*}
\mathbf{P}_{y,a,i} = e^{\mathbf{A}_{y,a,i}\delta_i}
\end{equation*}
The instantaneous rate matrix takes rates of movement between regions and the mortality rates for each fleet and region. Along the diagonal is the negative of the sum of the other rates (the hazard) so each row sums to zero. For two regions and one fleet operating in each region:
\begin{equation*}
 \mathbf{A}_{y,a,i} = \begin{bmatrix}
 a_{y,a,i,1} & \mu_{1\rightarrow 2,y,a,i} & F_{y,a,i,1} & 0 & M_{y,a,1} \\
 \mu_{2\rightarrow 1,y,a,i} &  a_{y,a,i,2} & 0 & F_{y,a,i,2} & M_{y,a,2} \\
 0 & 0 & 0 & 0 & 0 \\
 0 & 0 & 0 & 0 & 0 \\
 0 & 0 & 0 & 0 & 0
 \end{bmatrix}
\end{equation*}
where \(a_{y,a,i,r} = -(\mu_{r\rightarrow r',y,a,i} + F_{y,a,i,r_f} + M_{y,a,r})\). When there is one region, \(n_f\) fleets, and \(\delta_i = 1\), exponentiating the instantaneous rate matrix results in the PTM defined in Eq. \ref{eq:ptm_1_region}.

@@

When there are \(n_R\) regions and \(n_F\) fleets, the vector of abundance in each state at age \(a>1\) on January 1 is \(\mathbf{N}_{y,a} = (\mathbf{N}_{O,y,a}^T, \mathbf{0}^T)^T\) where \(\mathbf{N}_{O,y,a} = (N_{y,a,1}, \ldots, N_{y,a,n_R})^T\) is the number in the states corresponding to being alive in each region and \(\mathbf{0}\) is a vector (\(n_F+1\)) for the numbers captured in each fleet and dead from natural mortality because no age \(a\) fish have died yet on January 1.

Each stock \(s\) is assumed to spawn and recruit in one region \(r_s\). So for age \(a=1\), \(\mathbf{N}_{O,y,1}\) is 0 except for row \(r = r_s\). Options for configuring recruitment (\(N_{y,1,r_s}\)) for each stock are the same as previous versions of WHAM (Miller and Stock 2020). If recruitment is assumed to be a function of SSB, it is only the spawning population in region \(r_s\) at the time of spawning that constitutes the SSB in the stock-recruit function. However, models can configure spawning individuals to occur in other regions at the time of spawning. Aside from treating recruitment as a random walk, the general model for annual recruitment as random effects is
\begin{equation*}
\log\left(N_{y,1,r_s}\right)|\text{SSB}_{y-1,r_s} =  f\left(\text{SSB}_{y-1,r_s}\right) + \varepsilon_{y,1,r_s}
\end{equation*}
where\\
\[\text{SSB}_{y,r_s} =  \sum^A_{a=1}  w_{y,a} \text{mat}_{y,a} \mathbf{O}_{y,a,r_s}^T(t_s) \mathbf{N}_{O,y,a}\]
and \(w_{y,a}\) is the mean weight at age of spawning individuals, \(\text{mat}_{y,a}\) is the maturity at age, and \(\mathbf{O}_{y,a,r_s}(t_s)\) are the probabilities of surviving and occurring in region \(r_s\) at time \(t_s\) given being alive in each region at the start of the year (the \(r_s\) column of the upper-left submatrix of Eq. \ref{eq:ptm_spawn}).

As in previous versions of WHAM, the apparent survival (and movement) from one year to another after recruitment can be treated deterministically or as functions of random effects. The predicted numbers at age in year \(y\) at age \(a\) for a given stock are vector analogs (\(\mathbf{N}_{O,y,a}\)) of the equations for numbers at age in the standard WHAM model (Stock and Miller 2021). For ages \(a = 2,\ldots, A-1\), where \(A\) is the plus group, the expected number alive in each region at the beginning of the following year and next age class age can be obtained from the first \(n_R\) elements of the vector
\[\mathbf{P}_{y-1,a-1}^T \mathbf{N}_{y-1,a-1}.\]
The numbers alive in each region can also be modeled more simply using the sub-matrix \(\mathbf{O}_{y,a}\). The general model for the vector of abundance at age in year \(y\) at age \(a\) given the vector in year \(y-1\) at age \(a-1\) is
\begin{equation*}
\log\left(\mathbf{N}_{O,y,a}\right)|\mathbf{N}_{O,y-1,a-1} =  \log\left(\mathbf{O}_{y-1,a-1}^T \mathbf{N}_{O,y-1,a-1}\right) + \boldsymbol{\varepsilon}_{y,a}
\end{equation*}
for ages \(a = 2,\ldots, A-1\), and for the plus group
\begin{equation*}
\log\left(\mathbf{N}_{O,y,A}\right)|\mathbf{N}_{O,y-1,A-1},\mathbf{N}_{O,y-1,A} = \log\left(\mathbf{O}_{y-1,A-1}^T \mathbf{N}_{O,y-1,A-1} + \mathbf{O}_{y-1,A}^T \mathbf{N}_{O,y-1,A}\right) + \boldsymbol{\varepsilon}_{y,A}
\end{equation*}
where \(\boldsymbol{\varepsilon}_{y,a}\) is the vector of region-specific errors for a given stock, which are independent across stocks and regions, but the same correlation structures as previous versions are possible across ages and years for a given stock and region.
When there is autocorrelation with age, WHAM now assumes this applies only to ages \(a>1\) by default so that recruitment random effects are independent of those for the apparent survival of older age classes. So the general covariance structure for a given stock at ages \(a>1\) in region \(r\) is that of a two-dimensional first-order autoregressive (2DAR1) process
\begin{equation*}
  Cov\left(\varepsilon_{y,a,r},\varepsilon_{y',a',r}\right) =   \frac{\rho_{N,\text{age},r}^{|a-a'|}\rho_{N,\text{year},r}^{|y-y'|}\sigma_{N,a,r}\sigma_{N,a',r}}{\left(1 -  \rho_{N,\text{age},r}^2\right)\left(1 - \rho_{N,\text{year},r}^2\right)} 
\end{equation*}
and that for age 1 is just AR1 across years
\begin{equation*}
  Cov\left(\varepsilon_{y,1},\varepsilon_{y',1}\right) =   \frac{\rho_{N,1,\text{year}}^{|y-y'|}\sigma^2_{N,1}}{1 - \rho_{N,1,\text{year}}^2}.
\end{equation*}
When the annual changes in abundance at age are treated deterministically, \(\boldsymbol{\varepsilon}_{y,a} = 0\). Since recruitment for a given stock currently only occurs in one region \(r_s\) there is only a single vector of time-varying recruitment random effects for each stock.

@@

Initial numbers at age for each stock and region can be treated as age-specific fixed effects or with an equilibrium assumption as in previous versions of WHAM. For the equilibrium option there are two parameters for each stock: the stock-specific fully-selected fishing mortality rate \(\log \widetilde{F}\) and the recruitment in year 1 \(\log N_{1,1,r_s}\). A stock-specific equilibrium fishing mortality at age by fleet \({\tilde F}_{a,f}\) is the product of \(\widetilde{F}\) and the selectivity across fleets in the first year
\begin{equation*}
  \text{sel}_{1,a,f} = \frac{F_{1,a,f}}{\max_a \sum_{f=1}^{n_F} F_{1,a,f}}.
\end{equation*}
We use \(\widetilde{F}_{a,f}\) to define an equilibrium abundance per recruit by region at age \(a\) conditional on recruiting to each region
\begin{equation}\label{eq:npraa}
 \widetilde{\mathbf{O}}_{a} = \left\{
 \begin{array}{ll}
\prod^{a-1}_{j=0}\mathbf{O}_{j}  & 1\leq a<A\\
\left[\prod^{a-1}_{j=0}\mathbf{O}_{j}\right] \left(\mathbf{I} - \mathbf{O}_{A}\right)^{-1} & a = A
 \end{array}
\right.
\end{equation}
where \(\mathbf{O}_{j}\) is the equilibrium probability of surviving age \(a\) and occurring in each region and \(\mathbf{O}_{0} = \mathbf{I}\). Natural mortality and movement rates in the first year of the model are also used in Eq. \ref{eq:npraa}. For the plus group \(a=A\), \(\left(\mathbf{I} - \mathbf{O}_{A}\right)^{-1}\) is a ``fundamental matrix'\,' derived using the matrix version of the geometric series (Kemeny and Snell 1960). Recall that recruitment for stock \(s\) only occur in region \(r_s\) so, the equilibrium initial numbers at age \(a\) by region are
\[\mathbf{N}_{O,1,a} = \widetilde{\mathbf{O}}_{a}^T \mathbf{N}_{O,1,1}.\]

@@

When treated as constant parameters, natural mortality rates may be stock-, region-, and age-specific, but currently are constant across seasons. When random effects are used, the same 2DAR1 structure with age and year as described by Stock and Miller (2021) can be configured for a given stock and region and any environmental covariate effects can be stock-, region-, and age-specific. So the general model for natural mortality is
\begin{equation*}
  \log M_{y,a,r} = \theta_{M,r} + \varepsilon_{M,r,y,a} + \sum^{n_E}_{k=1} \beta_{M,r,a,k} E_{k,y}.
\end{equation*}
The general covariance structure for random effects are modeled most generally as 2DAR1 random effect with age and(or) year where the covariance is
\begin{equation*}
  Cov\left(\varepsilon_{M,y,a,r},\varepsilon_{M,y',a',r}\right) =   \frac{\rho_{M,\text{age},r}^{|a-a'|}\rho_{M,\text{year},r}^{|y-y'|}\sigma^2_{M,r}}{\left(1 -  \rho_{M,\text{age},r}^2\right)\left(1 - \rho_{M,\text{year},r}^2\right)}.
\end{equation*}

@@

{\centering \includegraphics[height=0.95\textheight]{selectivity_south_plot}

\caption{Selectivty for fleets and indices in the southern region.}\label{fig:selectivity-south}
\end{figure}
\pagebreak
